Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
draft of eval 2

\paragraph{Promises}
The fact that arguments are evaluated lazily, i.e. expression are packing
into promises and evaluated on first access, complicates the information
gathering.  For example, some promises remain un-evaluated, and it would be
erroneous to force them as they may cause side-effects that will affect the
programs execution.  To deal with unevaluated arguments, we make an initial
guess for each argument at fufnction entry.  If the promise is later forced,
we simply update the recorded type for the argument.

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are termed
missing. This occurs when a function was called with too few arguments and
no default value was specified for missing arguments.  We record a
\k{missing} type for such argument.  There are two obvious ways to deal with
missing arguments: type them as \any or type them as some unit type.  As we
are performing a dynamic analysis, we conservatively type them as \any.

\paragraph{Non-local returns}
When a function exits with a longjump there is no return value to speak
of. In order to ensure that call traces are valid when a longjump occurs, we
intercept the unwinding process initiated by the longjump and mimics the
functionality of the functions being exited. When no return value is present
for a call, we record a special \k{jumped} return type.

\paragraph{Implementation details.}
We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},
\k{special\_exit}, \k{promise\_force\_entry}, and
\k{prom\-ise\_force\_exit}.  The function-related callbacks are used mainly
for bookkeeping: the analysis is notified that a construct has been entered
by pushing the call onto a stack.  The calls themselves store a trace
object, it is that object that holds the type information. As R can perform
single or multiple dispatch on function arguments depending on their class,
the relevant information is kept by the {\tt \_entry} variants.

@@

\begin{tabular}{llrrr}
\toprule
Actual Type & Expected Type & \# Failed Assertions & \% Total & Cumulative \%\\
\midrule
double[] & class<`matrix`> & 371896 & 27.37 & 27.37\\
double[] & integer[] & 117212 & 8.62 & 35.99\\
character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
double & integer | null & 60570 & 4.46 & 47.81\\
double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
\addlinespace
class<matrix> & class<`timeSeries`> & 55005 & 4.05 & 56.19\\
double[] & double & 53338 & 3.92 & 60.12\\
double & class<`data.frame`> & 32553 & 2.40 & 62.51\\
double[] & class<`data.frame`> & 31576 & 2.32 & 64.84\\
double & integer & 20586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}
\caption{Top types of contract assertion failures.}
\end{table}

@@

Turning our attention now to arguments, we found that only \PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting runtime occurrences, but that data alone does not tell the full story, as some failures may be overrepresented if e.g. a failing contract assertion was in a loop.
We were interested in a breakdown of type signature by parameter, and seeing how often those contracts were violated.
Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by parameter position, folding away multiple identical failed contract assertions for the same parameter position.
The first row of this table reads: a value of type ${\bf chr}[]$ was passed to 183 different argument positions expecting a value of type ${\bf chr}$, representing 2.05\% of all arguments where a contract check failed.

The main takeaway from Table~\ref{tbl:fail-breakdown-1} is that there appear to be no pathological kinds of failing assertions: programmers are not disproportionately passing values of one type to arguments expecting values of another.
In other words, our type language appears to match with programmer intuitions: had we failed to account for a key, type-based assumption that R programmers make, we would see it at the top of the table.

@@

% END Autogenerated
 \label{tbl:fail-breakdown-1}
 \caption{Most common failing contract checks by type. Note that for this table, we collapsed multiple identical failed assertions into one, meaning that single arguments with multiple (identical) failed assertions are only counted once.}
 \end{table}

%
%
% \subsubsection{Patterns in Failed Contract Assertions}

% While Table~\ref{tbl:fail-breakdown-1} revealed no glaring patterns in failed assertions, there are a few subtle issues that we will discuss.
% First, \PERCFAILEDVECTOSCA of failed checks broken down by parameter position failed due to vectors being passed where scalars were expected.

% We observed an interesting corner case and quirk of R in looking at functions with values that default to NA.
% In R, recall that there is a different type of NA for each primitive, and when the literal value \code{NA} appears in code, R assumes it to be the logical NA.
% For the purposes of our analysis, this has the effect of introducing $\string^{\bf lgl}$ types where they (probably) don't belong, particularly when programmers fail to effectively test their code.
% For example, a function header like \code{function(x, y=NA)} will have $\string^{\bf lgl}$ as part of the type of parameter \code{y}.
% We tried to control for this in our analysis, and indeed $\string^{\bf lgl}$ is a subtype of all NA-able numeric types.

>>>>

\paragraph{Promises}

The goal of our analysis is to discover argument types for functions,
builtins, and specials (R functions such as \{, \code{if}, etc.).  First,
recall that R wraps arguments in promises (Section~\ref{sec:R}).  Naively
intercepting function calls and recording argument types would result in
incomplete data: While we can dig into promises without forcing them, if the
innermost value is an unevaluated expression we are unable to probe it
without evaluating it, which may alter program behaviour.  To circumvent
this obstacle, we instead intercept the forcing of promises: when a promise
is forced, we identify the surrounding call, collect the type of the value,
and add it to the call trace associated with the current function call.
Once function execution completes, we collect information about the return
value, add it to the call trace, and save the completed call trace to be
written out later.

First, we wanted to ensure that \typetracer captured types for unevaluated
arguments.  Unfortunately, not every argument of a function is necessarily
used during every function execution, meaning that some arguments may remain
unevaluated (as they are never forced), and there are a few reasons for
this.

One possibility is that the argument was simply unused given the control
flow of this particular execution (e.g. if the argument is only needed in
one branch of a conditional expression).  This mainly occurs when arguments
have default values for some parameters that are not critical to every
execution of the function.  In these cases, we still want to collect the
type of the default value.

Another possibility is that a value is used {\it without being forced}, as
is the case when values are ``metaprogrammed''.  This is common practice in
e.g. the \code{tidyverse}~\footnote{\AT{either talk about it here or in
    background}} ecosystem, as they use the metaprogramming capabilities of
R to implement a DSL for data analysis.

To tackle possibly unevaluated arguments, we make an {\it initial guess} of
function argument types on function entry. Specifically, we collect types
for argument default values, and ascribe an ``unused'' type for unevaluated
expressions.  If an argument promise is later forced, we simply update the
recorded type for the argument.

\paragraph{Missing arguments}
Second, we wanted to account for formal arguments (arguments which are named
in the function definition) which received no values when the function was
called. This represents cases where a function was not called with all of
its arguments, and no default value was specified for missing arguments.  We
record a ``missing'' type for such argument.

There are two obvious ways to deal with missing arguments: type them as $\bot$
or as $\top$.  As we are performing a dynamic analysis based on package test
code, we conservatively type these arguments as $\top$, or \code{any}, as It is
impossible to know dynamically what all possible function behaviours are.
We discuss the prevalence of such any-typed arguments in
Section~\ref{sec:evaluation}.

\paragraph{Non-local returns}
Third, we had to contend with R's backend sometimes performing longjumps,
which discards current R calls from the call stack.  In order to ensure that
call traces are not lost when a longjump occurs, \typetracer intercepts the
unwinding process initiated by the longjump and mimics the functionality of
the functions being exited.  The R-dyntrace framework supplies the return
value during the unwinding process (if there is indeed such a value), so we
can capture it, analyze it, and include it in the call trace.  In certain
situations, however, no return value can be obtained (e.g. a longjump
occurred as part of some error handling), and then we simply record a
``jumped'' return value, which we can deal with in post-processing.

\paragraph{Implementation details.}
We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},\k{special\_exit},
\k{promise\_force\_entry}, and \k{promise\_force\_exit}.  The
function-related callbacks are used mainly for bookkeeping: the analysis is
notified that a construct has been entered by pushing the call onto a stack.
The calls themselves store a trace object, it is that object that holds the
type information. As R can perform single or multiple dispatch on function
arguments depending on their class, the relevant information is kept by the
{\tt \_entry} variants.

The primary functionality of \typetracer is in {\tt promise\_force\_exit}.
As arguments to functions in R are wrapped in promises, we delay our
reflection to promise force time.  In R, promises are essentially objects
with value and expression slots.

When a promise is created for a particular expression, that expression is
merely put into the expression slot of a promise object, and a special {\it
  unbound} value is loaded into the value slot (unless some default value is
specified, wherein it is loaded into the value slot---this is the case with
default argument values).  When the promise is forced, the expression is
evaluated in the forcing context, and the value is stored in the value slot
of the promise.  So to get the type of a function argument, we check the
value slot of the argument promise---recursively if the value itself is a
promise.

@@

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
\midrule
double[] & class<`matrix`> & 371896 & 27.37 & 27.37\\
double[] & integer[] & 117212 & 8.62 & 35.99\\
character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
double & integer | null & 60578 & 4.46 & 47.81\\
double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
\addlinespace
class<matrix> & class<`timeSeries`> & 55005 & 4.05 & 56.19\\
double[] & double & 53338 & 3.92 & 60.12\\
double & class<`data.frame`> & 32553 & 2.40 & 62.51\\
double[] & class<`data.frame`> & 31576 & 2.32 & 64.84\\
double & integer & 20586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}
\caption{Top types of contract assertion failures.}
\end{table}

@@

Turning our attention now to arguments, we found that only \PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting runtime occurrences, but that data alone does not tell the full story, as some failures may be overrepresented if e.g. a failing contract assertion was in a loop.
We were interested in knowing for each of the most common violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments had that type, and how many of those exhibited the contract failure in question.
Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by unique parameter position, folding away multiple identical failed contract assertions for the same parameter position.
The first row of this table reads: a value of type \code{double[]} was passed to 15 different function parameters expecting a matrix, of which there are 1394 in total: 1.08\% of matrix-typed parameters were passed double vector values.

We see that even though the double vector and matrix issue was wildly prevalent in the raw, dynamic contract evaluation numbers, the number of actual function argument types that were violated is very small.
The story is similar with the double and integer coercion we mentioned earlier: it represents many dynamic contract failures, but very few of the integer vector-typed arguments have their contracts violated by double vector-typed values.
Rows 5 and 6 are interesting: we see that rather often arguments expecting time series data are passed double vector or matrix values.
This is a quirk of the \code{timeSeries} package, whose functions often accept matrices and vectors, converting them to time series in an ad hoc manner.
Note that the code coverage of the \code{timeSeries} tests, examples, and vignettes on the \code{timeSeries} package code is only 58.24\%, which is one possible explanation of why these contract failures are occurring: the types that \typetracer generates are only as good as the test code its run on.

@@

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
double[] & class<`matrix`> & 15 & 1394 & 1.08\\
double[] & integer[] & 44 & 657 & 6.70\\
character & class<`bignum`> | raw[] & 1 & 3 & 33.33\\
double & integer | null & 5 & 20 & 25.00\\
double[] & class<`timeSeries`> & 19 & 35 & 54.29\\
\addlinespace
class<matrix> & class<`timeSeries`> & 19 & 35 & 54.29\\
double[] & double & 100 & 5021 & 1.99\\
double & class<`data.frame`> & 1 & 900 & 0.11\\
double[] & class<`data.frame`> & 6 & 900 & 0.67\\
double & integer & 48 & 529 & 9.07\\
\bottomrule
\end{tabular}

% END Autogenerated
 \label{tbl:fail-breakdown-1}
 \caption{Results from Table~\ref{tbl:fail-breakdown-0}, broken down by occurrences of the expected type as a parameter type.}
 \end{table}

Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently violated contracts amongst the most frequently occurring argument types.
We selected argument types which were in the 90th percentile of argument type occurrences, computed the most frequent type signature violations among them, and reported the most frequently violated contracts together with the type of the value that violated that contract. 
The first row of the table reads: 44 function arguments with integer vector type are passed double vectors instead, and 657 arguments have integer vector type.
Had we failed to capture some key usage pattern of R with our type annotation framework, we would likely see it here, and we can see this in action if we consider Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile instead.
The most frequent argument type violation pattern in that of data frames being passed to arguments expecting tibbles: this occurs in only 22\% of such arguments, and represents cases where tests did not adequately cover all valid function inputs.
Separate from the issue of testing, we can capture this behaviour with user-defined subtyping, as data frames are subtypes of tibbles, but this necessitates a proper object-oriented type system for R which is outside the scope of this work.

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
double[] & integer[] & 44 & 657 & 6.70\\
\textasciicircum{}logical[] & null & 158 & 2756 & 5.73\\
character[] & character & 183 & 4102 & 4.46\\
character & null & 117 & 2756 & 4.25\\
double & null & 107 & 2756 & 3.88\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-2}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile.}
 \end{table}

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
class<data.frame> & class<`data.frame`, `tbl`, `tbl\_df`> & 63 & 284 & 22.18\\
class<matrix> & class<`array`> & 13 & 70 & 18.57\\
\textasciicircum{}logical[] & logical | null & 9 & 49 & 18.37\\
character[] & list<character[]> & 20 & 109 & 18.35\\
complex[] & complex & 7 & 42 & 16.67\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-3}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.}
 \end{table}

In sum, we believe that this evaluation shows that the type signatures we generate from traces are quite good.
Only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed at runtime, representing failures in as few as \PROPARGSFAILINGASSERTS of argument types.
Even though \PROPFUNSFAILEDCHECKNOSTHREE of functions had at least one argument type involved in a failing contract check, these functions are under-tested, representing only \PERCCALLSBADFUNSINTESTS of calls observed while inferring types.
